Preparing for Potential Closure of European Airspaces due to Reentering Space Objects

### 2025-04-30 paper.qmd
semi final?

---
title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"

@@

# Abstract
The risk of collision of re-entering space debris with aircraft is 
increasing due to both increasing number of airborne flights and 
re-entries. National and international authorities need to balance
safety and economic impact of adopted measures.  
We asses airspaces density as flown by commercial airlines
distinguishing seasonal and hourly patterns.
We base our hourly density estimation on (enhanced) flown trajectories as
recorded by EUROCONTROL's \acr{NM} binned over Uber's H3
gridding system. Impact risk probabilities will then calculated based on work
from [@patera2008].
This combined with a buffered area along the the path of re-entry,
can provide risk impact probabilities for deciding on safety measure such as
airspace closures and/or re-routings.

@@

We base our hourly flight density estimation on trajectories flown
by \acr{IFR} commercial aircraft as
recorded by EUROCONTROL's \acr{NM}[^1] and overlapped on Uber's H3
gridding system at different resolutions [@h3resolutions].
@fig-cells shows cells for resolution $2$ which corresponds to hexagons
of average area $\qty{86801.78}{km^2}$ and average edge length of 
$\qty{182.51}{km}$, see also @tbl-resolutions for other resolutions.

```{r}
#| label: tbl-resolutions
#| tbl-cap: H3 hexagons, average values
#| tbl-cap-location: bottom

library(tidyverse)
library(gt)
library(gtExtras)

tribble(
  ~resolution, ~hex_area, ~hex_edge,
  2L , 86801.780398997, 182.5129565,
  3L,  12393.434655088,  68.97922179,
  4L,   1770.347654491,  26.07175968
) |> 
  gt() |> 
   cols_label(
    resolution ~ "Resolution",
    hex_area ~ md("Cell Area"),
    hex_edge ~ md("Edge length")
  ) |> 
    gt::tab_header(
    title = md("H3 Cell Statistics"),
    subtitle = md("average hexagons")
  ) |> 
  cols_units(
    hex_area = "km^2",
    hex_edge = "km"
  ) |>
  opt_align_table_header(align = "left")
```

@@

```{r cells-map}
#| out-width: "100%"
#| echo: false
#| fig-scap: "H3 cells at resolution 2"
knitr::include_graphics("media/figures/cells_2.png")
```

H3 cells (red) at resolution 2 and their centers (blue) covering EUROCONTROL Area.
:::

We characterise densities on an hourly basis for representative days
in summer and winter.[^2]

We expand on the works of [@wrightAirspaceClosuresDue2025; @patera2008; @ailorLargeConstellationDisposal2020] and
provide a reproducible example for States, Agencies or relevant authorities
to build upon.

@@

[^2]: We will consider representative days in \acr{IATA} summer 2024 
and winter 2025 season.

@fig-density-winter-res-3 provides a promising result in the hourly flight
density calculation which at least shows
expected outcomes such as cells at zero density for war or no-fly zones such as
Ukraine, Libia or Syria.

::: {#fig-density-winter-res-3}

```{r density-winter-res-3}
#| out-width: "110%"
#| echo: false
knitr::include_graphics("media/figures/flight_density_2024-08-01T12_3.png")
```

Flight density in the EUROCONTROL Area on 2024-08-01T12 (H3 cells at resolution $3$).
:::

The probability of impact for debris object $k$, $p_k$, is expressed as.

$$
p_k = \varepsilon_k \int_a \rho(\vec{x}) \sigma(\vec{x}) da
$$ {#eq-impact-probability}

where $\rho({\vec{x}})$ is the probability density of an aircraft
and $\sigma(\vec{x})$ is probability density of a debris object in the
position $\vec{x}$.

## Impact probability density

Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to reentry even if initially on an elliptical ones.
In fact drag at perigee will lower apogee and similarly drag at apogee will
lower perigee. But because drag at perigee is higher due to denser atmosphere,
the apogee  will decrease faster than perigee and the orbit will then tend to be
a circular one.
The angular position of the reentry point from the ascending node, 
see @sec-orbitals, is uniformly distributed, $\sigma(\theta) = 1 / (2\pi)$,
due to uncertainties in atmospheric drag and the gravitational effects of
the Equatorial bulge, hence we assume that the impact longitude, $\lambda$,
is uniformely distributed $\sigma(\lambda) = 1/(2\pi)$.

@@

So for example [@patera2008], if a building is located at latitude 30 degrees
north (or south) and has an area of $\qty{1000}{m^2}$ and a piece of debris
reenters the Earth’s atmosphere from an orbit having an inclination of 45 degrees,
the impact probability is given by

@@

$\num{0.046}$

## Aircraft effective exposed area
From Patera

> Even for a ballistic coefficient of 50 lbs/ft2, the flight path angle
> of a reentering object at 40,000 feet is -86.6 degrees, which corresponds
> to nearly vertical descent.
> Since most debris objects have ballistic coefficients equal to or 
> less than 50 lbs/ft2, the vertical descent assumption is a very
> accurate approximation.

The effective expose area, $\varepsilon_i$, is
$$
\varepsilon_i = \frac{s_{a} a_{h} + s_{d} a_{v}}{s_d}
$$
where $s_{a}$ is the average aircraft speed, $s_{d}$ is the average debris 
object fall speed, $a_v$ area of the aircraft as viewed from above,
$a_h$ area of the aircraft as viewed from the from the front of the aircraft.

# Data

>>>>

---
title: "Preparing for Potential Closure of European Airspaces due to Reentering Space Objects"

@@

# Abstract
The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
National and international authorities need to balance safety and economic
impact of adopted measures should a reentry event be announced, usually on short
notice.  
We employ hourly and daily spatial density of flight calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned over Uber's H3 gridding system.  
Impact risk probabilities are then calculated based on previous theoretical
work but using these flights trajectories instead of proxies such as
population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide risk impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

The ever-growing forecast number of flights, see @fig-forecast, and the modern
race to commercialisation of space call for a reevaluation of the risk of
collision and casuality due to uncontrolled debris reentries.
Oversighting authorities need methodologies and tools to assess which 
safety measures are the most adequate to minimize casualties and economic
impact.

Surging number of flights in Europe [@forecastUpdate2025] and in the
world, but more importantly under development/deployment megaconstellations
of satellites on \acr{LEO} such as already active Starlink and OneWeb or
planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
of orbiting objects with a short operational life some of which will 
inevitably reenter Earth atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].

This paper develops a practical and straightforward pipeline to rapidly
assess the risk from uncontrolled reentry of space debris for the European
region.
Apart from the general risk assessment of potential reentries, we analyze
a soon to happen reentry of debris from a 1972 failed USSR mission to Venus.

::: {#fig-forecast}

```{r forecast}
#| out-width: "100%"
#| echo: false
#| fig-scap: "Forecast"
knitr::include_graphics("media/figures/forecast_statfor.png")
```

EUROCONTROL 7-year forecast 2025-2031 (Spring 2025).
:::

```{r}
#| label: tbl-megaconstellations
#| tbl-cap: Megaconstellations (planned > 1000) as per filing to the \acr{ITU}
#| tbl-cap-location: bottom

library(tibble)
library(gt)

megas <- tibble::tribble(
  ~constellation, ~state, ~launched, ~active, ~planned, ~first_launch,
  "Starlink", "US", 4714, 3521, 4714, 2018,
  "Starlink2A", "US", 3689, 3070, 6720, 2022,
  "Starlink2", "US", 0, 0, 30456, NA,
  "OneWeb", "UK", 660, 635, 716, 2019,
  "OneWeb2", "UK", 0, 0, 2304, NA,
  "Kuiper", "US", 29, 27, 3232, 2023,
  "StarShield", "US", 193, 126, 32, 2022,
  "Xingwang", "CN", 50, 10, 996, 2021,
  "Qianfan", "CN", 90, 28, 32, 2024,
  "Guangwang", "CN", 0, 0, 12992, NA,
  "Yinhe", "CN", 8, 7, 1000, 2020,
  "Hanwha", "KR", 0, 0, 2000, NA,
  "Lynk", "US", 10, 6, 2000, 2020,
  "Astra", "US", 0, 0, 13620, NA,
  "Telesat", "CA", 0, 0, 300, NA,
  "HVNET", "US", 0, 0, 1440, NA,
  "SpinLaunch", "US", 0, 0, 1190, NA,
  "Globalstar3", "DE", 0, 0, 3080, NA,
  "Honghu-3", "CN", 0, 0, 10000, NA,
  "Semaphore", "FR", 0, 0, 116640, NA,
  "E-Space", "US", 4, 0, 337323, 2022
)

# ~constellation, ~state, ~launched, ~active, ~planned, ~first_launch,
megas |> 
  gt() |> 
   cols_label(
    constellation ~ "Constellation",
    launched ~ "Launched",
    active ~ "Active",
    planned ~ "Planned",
    first_launch ~ "First launch"
  ) |> 
   sub_missing(
    columns = "first_launch",
    missing_text = "-"
  ) |> 
  opt_table_font(size = "10px")

```

We follow the same mathematical approach of 
[@patera2008; @wrightAirspaceClosuresDue2025] but instead of
approximating flight density with population density [@patera2008] or
aircraft transponder data [@wrightAirspaceClosuresDue2025], we use flown 
trajectories.

We base our hourly flight density estimation on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1] and binned on Uber's H3 gridding system
[@uber-h3] for fast processing.  
This work will use H3 cells at resolution 3 which corresponds to hexagons of
average area $\qty{12393}{km^2}$ and average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

@@

```{r cells-map}
#| out-width: "120%"
#| echo: false
#| fig-scap: "H3 cells at resolution 3"
knitr::include_graphics("media/figures/cells_3.png")
```

H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
box of 43 EUROCONTROL Member States FIR's (blue shaded polygon).

We characterise densities on an hourly basis for the days of the 2023 and 2024
busiest \acr{IATA} summer ISO week.

We expand on the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
provide a reproducible example for States, Agencies or relevant authorities
to build upon.

@@

Next sections will develop the mathematcal background for the evaluation of
collision and casuality risk of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for 2024-07-05.
Expected outcomes such as cells at zero density for war or no-fly zones
(Ukraine, Libia or Syria) or typical traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted.

::: {#fig-density-summer2024-res-3}

```{r density-winter-res-3}
#| out-width: "110%"
#| echo: false
knitr::include_graphics("media/figures/flight_density_2024-07-05_3.png")
```

Flight density in the EUROCONTROL Area on 2024-07-05, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
:::

## Impact Probability for Single Flight, Single Debris
The probability, $p_k$, of a debris (point) object $k$, impacting an
aircraft (point) object over the area, $a$, is expressed as:

$$
p_k = \varepsilon_k \int_a \rho(\vec{x}) \sigma_k(\vec{x}) da
$$

where $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
$\sigma(\vec{x})$ is the impact probability density of a debris object $k$
in the position $\vec{x}$ and $\varepsilon_k$ is the effective exposed area,
i.e. the effective risk area for the $k^{th}$ debris object impacting
the aircraft (see @sec-eea later on).

If we discretaze the area of interest $a$ using H3 cells,
we have

$$
p_k = \varepsilon_k \sum_h \rho(h) \sigma_k(h)
$$

So for cell $h$ we have

$$
p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h)
$$

## Aircraft effective exposed area{#sec-eea}

We assume the aircraft impact area to be much larger than the debris
object area so as to neglet the debris impact area.
We can also simplify the geometry of the aircraft and model it as
a rectangular solid with dimesions of wingspan, length and height. 
Now, if the aircraft were stationary, the area exposed to a debris impact would
be the top view, while if the debris were stationary it would be the front view.
In reality the aircraft is moving (manly) horizontally and the debris 
vertically [^balistic_descent], which which allows to express the effective
exposed area, $\varepsilon_k$, as

$$
\varepsilon_k = \frac{s_{a} a_{h} + s_{d_k} a_{v}}{s_{d_k}}
$$

where $s_{a}$ is the average aircraft speed, $s_{d_k}$ is the average fall speed
of the $k$-th debris object, $a_T$ area of the aircraft as viewed from the top,
$a_F$ area of the aircraft as viewed from the from the front of the aircraft.

In our calculations we will use ($t$ subscript for aircraft type)

* constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$
* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, 
* aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$, 
  where $W_t$ is the wingspan, and $L_t$ the length for the specific aircraft
  type $t$
* aircraft area as viewed from the front, $a_F = W_t \times H_t = F_t$, 
  where $W_t$ is the wingspan, and $H_t$ the height for the specific aircraft
  type $t$

The effective exposed area now only depends on the aircraft type $t$ (dimensions
and cruise speed) and can be rewritten as:

$$
\varepsilon_t = \frac{C_{t} F_{t} + S A_{t}}{S}
$$

The values $W_t$, $H_t$ and $L_t$ have been retrieved from [@performance_database;@skybrary;@doc8643].

The impact  probability for cell $h$ and single aircraft of type $t$ becomes:

$$
p_{t,k}(h) = \varepsilon_t \rho_t(h) \sigma_k(h)
$$

[^balistic_descent]: *Even for a ballistic coefficient of 50 lbs/ft2, the flight
path angle of a reentering object at 40,000 feet is -86.6 degrees, which
corresponds to nearly vertical descent.
Since most debris objects have ballistic coefficients equal to or 
less than 50 lbs/ft2, the vertical descent assumption is a very
accurate approximation.* [@patera2008]

## Impact probability density for debris of inclination $i$

Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to reentry even if initially on elliptical
ones.[^circular-orbit]

The angular position, $\theta$, of the reentry point from the ascending node, 
see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
$\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in atmospheric drag
and the gravitational effects of the Equatorial bulge.
Hence we assume that the impact longitude, $\lambda$, is uniformely distributed.

@@

where $R$ is the distance from the center of the Earth of the object potentially
being impacted by the debris.

So for example if a building is located at latitude 30 degrees and has an area 
of $\qty{1000}{m^2}$ and a piece of debris reenters the Earth’s atmosphere
from an orbit having an inclination of 45 degrees, the impact probability is 
given by

@@

The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic
U shape.

::: {#fig-weighting-inclination-i}

```{r weighting-inclination-i}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting_inclination_i.png")
```

Impact probability density $\sigma(\varphi, i = 51.7)$ for debris of inclination $i$ and altitude $\qty{550}{km}$. 
:::

[^circular-orbit]: In fact drag at perigee will lower apogee and similarly drag
at apogee will lower perigee. But because drag at perigee is higher due to
denser atmosphere, the apogee  will decrease faster than perigee and the orbit
will then tend to be a circular one.

## Impact probability density for debris

If we consider historical debris reentry data, we can numerically derive
the debris impact probability as a weighting function by summing up and averaging
all single densities:

$$ \sigma(\varphi) = \sum_k \sigma(\varphi, i_k)$$

From [@gcat] we extracted 1968 debris reentries during 25 calendar years from 
1st Jan 2000 till 31st Dec 2024.
The result of combining all the U-shaped densities can be seen in @fig-weighting.

::: {#fig-weighting}

```{r weighting}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting.png")
```

Weighting function derived from 25 years of reentries as an
empirical approximation of $\sigma(\varphi)$. 
:::

Spatially combining $\sigma(\lambda,\varphi)$ for the H3 cells in the
\acr{NM} area we obtain the weightings per cell (per hour) as in
@fig-weighting-nm-area.

::: {#fig-weighting-nm-area}

```{r weighting-nm-area}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting_on_Earth.png")
```

Weightings for H3 cells in \acr{NM} area. 
:::

## Hourly impact probability per H3 cell
By combining the hourly aircraft density (per aircraft type) with the relevant
weightings per H3 cell, we can numerically compute the hourly impact probability
for each hexagonal cell as:

$$
 E_h =  \sum_t W(h) \cdot \rho_{t}(h) \cdot \varepsilon_t
$$

where $E_h$ is the collision expectation in the hexagon $h$,
$W(h)$ is the median value of the weighting function over the area of hexagon $h$, $t$ is the \acr{ICAO} aircraft type,
$\rho_{t}(h)$ is the traffic density for type $t$ in cell $h$ and
$\varepsilon_T$ is effective exposed area for aircraft of type $t$.

::: {#fig-collision}

```{r collision}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/cubes.png")
```

Collision risk probability. 
:::

## Use Case: KOSMOS 482 Descent Craft reentry

In few days (at the time of writing this paper) around May 10, 2025, the
KOSMOS 482 Descent Craft should reenter the Earth atmosphere [@kosmos482reentry].
This is supposedly the lander module from a 1972 failed Soviet Venera mission
[@SpaceReviewKosmos] of \qty{480}{kg} weight and circa \qty{1}{m} diameter.
Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \qty{467}{degreesCelsius} (\qty{872}{degreesFarenheit})).
So it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

# Data and Code for Reproducibility

We have calculated collision and casuality risk probabilities for the two
busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
as logically depicted in @fig-data-densities is available in the companion
GitHub repository of the paper at <https://github.com/espinielli/flight_density_space_debris>.
We have also developed a companion R package `aviodebris` to heavylift the
various data processing steps to produce the results in the paper;
it is available at <https://github.com/espinielli/aviodebris>.
